# Reproducible Research: Peer Assessment 1

Project information can be found in the README.md in the same directory as this file.

```{r "Set Knitr options"}
opts_chunk$set(echo=TRUE) # technically not necessary since the default is TRUE
```

Let's initialize with the packages required for the analysis
```{r "Initialization"}
rm(list=ls()) # clean up environment when not using knitr to reduce side effects
library(ggplot2)
library(lubridate)
library(data.table)
```

## Loading and preprocessing the data
Provided activity.zip file in the repository is read an converted to a data.table
```{r "Loading and preprocessing"}
zipfile <- "activity.zip"
activityFileName <- "activity.csv"

activities <- read.csv(unz(zipfile, activityFileName))
activities$date <- ymd(activities$date) # reclassify factor as date

activities.dt <- as.data.table(activities)
```

Overview of the data.table we just read:
```{r}
summary(activities.dt)
str(activities.dt)
head(activities.dt)
```

## What is mean total number of steps taken per day?
First we need to calculate the total number of steps per day.
```{r "steps per day"}
totalStepsPerDay <- activities.dt[,list(totalSteps=sum(steps, na.rm=TRUE)), by=date]
summary(totalStepsPerDay)

g <- ggplot(totalStepsPerDay, aes(x=totalSteps))
g <- g + geom_histogram(binwidth=2000, color="lightgray")
g <- g + labs(title= "Histogram Of The Total Number Of Steps Per Day")
g <- g + labs(x= "Total Number Of Steps Per Day")
g + geom_rug() 

# mean and median total steps per day
totalStepsPerDay[,list(
    mean=mean(totalSteps), 
    median=median(totalSteps))]
```

## What is the average daily activity pattern?
In this section we are looking at the data summarized at the interval level
```{r "Average daily activity"}
avgStepsPerInterval <- activities.dt[,list(averageSteps=mean(steps,na.rm=TRUE)), by=interval]
head(avgStepsPerInterval)
```

```{r "plot of average steps per interval"}
g <- ggplot(avgStepsPerInterval, aes(interval, averageSteps))
g <- g + geom_line()
g + labs(title="Average Daily Activity Pattern", y = "Number of Steps")
```

We can now find the interval with the maximum average number of steps
```{r}
maxAverageNumberOfSteps <- max(avgStepsPerInterval$averageSteps)
result <- avgStepsPerInterval[averageSteps == maxAverageNumberOfSteps,interval]
```
The interval with the biggest average number of step is **`r result`**

## Imputing missing values
The number of missing steps in the dataset is : **`r nrow(activities.dt[is.na(steps)])`**

### Filling missing values
Strategy being used:  
Take the previously calculated mean number of steps for a given interval to replace the missing values.
```{r "Filling missing values"}
nomissing.dt <- activities.dt
nomissing.dt[is.na(steps), steps := as.integer(avgStepsPerInterval[avgStepsPerInterval$interval == interval,averageSteps])]
```

Let's now create the same histogram from above with the missing values filled.
```{r "Steps per day no NA"}
totalStepsPerDay.noNA <- nomissing.dt[,list(totalSteps=sum(steps, na.rm=TRUE)), by=date]
summary(totalStepsPerDay.noNA)

g <- ggplot(totalStepsPerDay.noNA, aes(x=totalSteps))
g <- g + geom_histogram(binwidth=2000, color="lightgray")
g <- g + labs(title= "Histogram Of The Total Number Of Steps Per Day (Without NA)")
g <- g + labs(x= "Total Number Of Steps Per Day")
g + geom_rug() 

# mean and median total steps per day without missing values
totalStepsPerDay.noNA[,list(
    mean=mean(totalSteps), 
    median=median(totalSteps))]
```

It appears that filling the missing values have removed a lot of the `0` and moved the mean and median to the right. 

## Are there differences in activity patterns between weekdays and weekends?
First we create our WeekDayType factor to be added to the data.table without missing values. Then the mean is calculated for each WeekDayType - Interval combination. 
```{r "prepare data.table"}
getWeekDayType <- function(date){
    ifelse(wday(date) %in% c(1,7),"weekday", "weekend")
}

nomissing.dt[, weekDayType := as.factor(getWeekDayType(date))]

nomissing.avgSteps<- nomissing.dt[,list(averageSteps=mean(steps,na.rm=TRUE)), by=list(weekDayType,interval)]

g <- ggplot(nomissing.avgSteps, aes(interval, averageSteps))
g <- g + geom_line()
g <- g + labs(title="Differences in Daily Activity Between Weekdays and Weekends")
g <- g + labs(y="Number of Steps")
g + facet_grid(weekDayType ~ .)
```

